Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-08-03 analysis/paper/paper.Rmd
Extended, results, discussion and conclusion

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Institute of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how it can potentially be used to understand cultural historical development in the region through the Mesolithic, c. 9500--4000 BCE.

@@

```{r shoredate, echo = FALSE, fig.cap = "A) An example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.  B) The geologically derived shoreline displacement curves for the region and the curve interpolated to the example site. C) The black line on the y-axis represents the elevation of the site. This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea, giving a \\textit{terminus post quem} date. D) The shoreline dating approach used for this study, where the gamma distrbution on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected. As is also indicated by the BIC values, this model represents the best explanation of the data of the compared models, while the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

@@

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development. Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded here.

@@

<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the models fit to the SPDs.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/bic.png"))
```

@@

Although there are generally few ^14^C-dates from the earliest part of the Mesolithic in Norway [e.g. @åstveit2018; @kleppe2018; @breivik2018a; @solheim2018], it is established that Stone Age sites in southern Norway date to the period from before the earliest ^14^C-dates in the study area for this paper, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8500 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022]. A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of a more forested landscape, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf. @damm2022; @sørensen2014]. While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.

The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. Following from what is generally held to be a reduction in residential mobility from the earliest to the later parts of the Mesolithic in the region [e.g. @lindblom1984; @fuglestvedt2018, 16--17], an immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.

@@

Following this earliest phase of rapid increase in shoreline dated sites, there a drop from this initial peak, starting from around 8800 BCE and concluding around 8000 BCE. It has previously been suggested that a change in mobility patterns occurs from some time around 8500--8300 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that has plateaued in the first centuries after 8000 BCE. The plateau of the logistic growth in the RSPD also has a corresponding positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.

@@

Subsequently, the SSPD deviates negatively from the exponential null from c. 6800 to c. 5800 BCE (a decrease also noted by @berg-hansen2022). This is followed by a positive deviation in the last centuries before 5000 BCE. The negative deviation could possibly been seen in relation to an increase in inland activity in southern Norway, which is to take place in the period from c. 6700--5900 BCE and reach its peak around 6200--6000 BCE [@boaz1999; @selsing2010; @persson2018]. This period is followed by a marked reduction in inland activity, which thus corresponds to the positive deviation in the SSPD, until a new increase occurs around 4500--4000 BCE [@persson2018, 204]. @boaz1999 related the first increase inland activity to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in coastal adaptation. The subsequent decrease in inland activity has been related to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022]. The Nøstvet period is argued to be characterised by a specialised coastal adaptation that also coincide with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in south-eastern Norway, starting from around [@fuglestvedt2018, 42--46; @glorstad2010, 216--233]. The onset of the classic Nøstvet phase thus roughly corresponds with the positive deviation in the SSPD around the later parts of the 6th millennium BCE.

Relating these patterns to demographic developments or land-use and mobility patterns is challenging given the data available here. Given the stationarity of the logistic model fit to the RSPD after c. 7500 BCE, is it possible that fluctuatons in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments. There appears to be a negative deviation from the logistic null model in the RSPD which corresponds to the one in the SSPD just before 6000 BCE. However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, given the present data. While it is possible that the fairly limited sample of ^14^C-dates might lead to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018].

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns. This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 7th millennium BCE. However, properly adjudicating this relationship and the degree to which it has been the same throughout the period, will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.

Another relevant point to mention here is that the relative sea-level change is effectively built into the Monte Carlo simulations. The critical envelopes resulting from the simulations therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the trajectory of relative sea-level change and a relationship between sites and shoreline that can constantly be described by the employed gamma distribution in the dating procedure. Beyond uncertainties to do with the dating procedure itself, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models. An increased movement instantiated by increased rates of relative sea-level change could result in the transition to a settlement pattern involving more frequent moves than that expected under a constant adjustment to the shoreline. If this is the case, the proxy would still reflect degree of occupational duration, but this could be a response to relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments. Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the cultural historical implications of variation in this proxy.

@@

This paper has presented a new methodology for assembling and assessing the result of summing a large number of shoreline dates, drawing on the probabilistic method for shoreline dating presented in @roalkvam2023, and analogous procedures for the treatment of radiocarbon dates [@crema2022; @timpson2014; @crema2021]. The result of this procedure was compared with the SPD of radiocarbon dates from within the same area to begin to assess the relationship between the two measures and their potential association with overall developments in population density and mobility patterns.

>>>>

---
title: "Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway"
# author:
#   - Isak Roalkvam, University of Oslo, Institute of Archaeology, Conservation and History
#   - Steinar Solheim, University of Oslo, Museum of Cultural History
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [iakh]
      correspondence: true
  - name: Steinar Solheim
    email: steinar.solheim@khm.uio.no
    institute: [mch]
    correspondence: false
institute:
  - iakh: University of Oslo, Institute of Archaeology, Conservation and History
  - mch: University of Oslo, Museum of Cultural History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: true
      toc: false
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - '--lua-filter=../templates/scholarly-metadata.lua'
      - '--lua-filter=../templates/author-info-blocks.lua'
      - '--lua-filter=../templates/pagebreak.lua'
header-includes:
  \usepackage{placeins}
  \usepackage[left]{lineno}
  \linenumbers
# always_allow_html: true
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how it can potentially be used to understand cultural historical development in the region through the Mesolithic, c. 9500--4000 BCE.

@@

```{r shoredate, echo = FALSE, fig.cap = "A) An example site (Askeladden ID 111967-1) relative to the isobases of the displacement curves.  B) The geologically derived shoreline displacement curves for the region and the curve interpolated to the example site. C) The black line on the y-axis represents the elevation of the site. This crosses the interpolated displacement curve at 7680 and 7490 BCE, which thus in practice gives an estimate of when the location of the site emerged from the sea, giving a \\textit{terminus post quem} date. D) The shoreline dating approach used for this study, where the gamma distrbiution on the y-axis describes the likely vertical relationship between the shoreline and the site when it was in use.",  out.width = "500px", fig.align="center", regfloat = TRUE}

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
The results of the Monte Carlo simulations are given in Figure \@ref(fig:spdmc), displaying the deviations from the simulation envelopes and the corresponding global p-values. The comparison of the models based on BIC scores is given in Figure \@ref(fig:bic).

For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected (Figure \@ref(fig:spdmc)E). As is also indicated by the BIC values, this model represents the best explanation of the data of the compared models, while the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD, around 8500 BCE, which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

@@

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A). Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded here.

@@

<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
```{r bic, echo = FALSE, fig.cap = "Bayesian information criterion for the models fit to the SPDs. Lower values indicate a better model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/bic.png"))
```

@@

Although there are generally few ^14^C-dates from the earliest part of the Mesolithic in Norway [e.g. @åstveit2018; @kleppe2018; @breivik2018a; @solheim2018], it is established that Stone Age sites in southern Norway date to the period from before the earliest ^14^C-dates in the study area for this paper, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8500 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022]. A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of forested landscape, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf. @damm2022; @sørensen2014]. While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.

The considered models also indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. Following from what is generally held to be a reduction in residential mobility from the earliest to the later parts of the Mesolithic [e.g. @bang-andersen2005; @fuglestvedt2018, 16--17; @lindblom1984; @bjerck2008, 103--105], an immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.

@@

Following this earliest phase of rapid increase in shoreline dated sites, there a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c. 7900 BCE. It has previously been suggested that a change in mobility patterns occurs from some time around 8500--8300 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE. The plateau of the logistic growth in the RSPD also has a corresponding positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.

@@

Subsequently, the SSPD deviates negatively from the exponential null from c. 6800 to c. 5800 BCE. This is followed by a positive deviation in the last centuries before 5000 BCE. These developments should possibly been seen in relation to simultaneous developments in the inland areas of southern Norway. The negative deviation in the SSPD appears to correspond to an increase in inland activity, which is to take place in the period from c. 6700--5900 BCE and reach its peak around 6200--6000 BCE [@boaz1999; @selsing2010; @persson2018]. @boaz1999 related this to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in coastal adaptation. This period is followed by a marked reduction in inland activity, lasting until c. 4500--4000 BCE [@persson2018, 204], the onset of which thus corresponds to the end of the negative deviation in the SSPD at c. 5800 BCE and the positive deviation from c. 5300--5000 BCE. @boaz1999 associated the period of decrease in inland activity to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022]. The Nøstvet period is argued to be characterised by an increase in specialised coastal adaptation that also coincide with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in Norway, starting from around 5700 BCE in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233]. The onset of the classic Nøstvet thus roughly corresponds with the positive deviation in the SSPD around the later parts of the 6th millennium BCE.

Directly relating these patterns to demographic developments or land-use and mobility patterns is challenging given the data available here. Given the stationarity of the logistic model of the RSPD after c. 7500 BCE, it is possible that fluctuations in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments. There appears to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE. However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data. While it is possible that the fairly limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018].

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns. This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 7th millennium BCE. However, properly adjudicating this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.

Another relevant point to mention here is that the relative sea-level change is effectively built into the Monte Carlo simulations, much in the same way as the Monte Carlo simulation method used with radiocarbon dates account for fluctuations in the calibration curve [e.g. @crema2022]. The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of relative sea-level change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma distribution employed for the dating procedure (Figure \@ref(fig:shoredate)D). Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models. An increased movement instantiated by increased rates of relative sea-level change could result in the transition to a settlement pattern involving more frequent moves than that which could be expected under a constant adjustment to the shoreline. If this is the case, the proxy would still reflect degree of occupational duration, but this could be a response to the rate of relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments. Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of variation in this proxy.

@@

This paper has presented a new methodology for assembling and assessing the sum of a large number of shoreline dates. This was done by employing the probabilistic method for shoreline dating presented in @roalkvam2023, and by drawing on analogous procedures in the treatment of the temporal frequency distribution of radiocarbon dates [@crema2022; @timpson2014; @crema2021]. The result of employing this methodology was then compared to radiocarbon dates from within the same area, to begin to assess the relationship between the two data types and their potential association with overall developments in population density and mobility patterns.

The main results were that the logistic model. This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from around 7500 BCE. This is contrasted to the shoreline dates, the development of which were better described by a function of exponential decay. While this indicates that some process of decline in the number of shoreline dated has occurred over time, the model was rejected as an adequate description of the data. Some of the deviations from the model were then narratively explored to suggest some possible reasons for the nature and reasons for these deviations, but given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions for future lines of interrogation.
